Lattices of acyclic pipe dreams

### 2023-04-26 acyclicFacets/draft/latticePipeDreams.tex
minor corrections

@@

We then explore in \cref{sec:furtherTopics} some natural further topics on pipe dreams.
We describe two algorithms to compute the unique acyclic pipe dream whose linear extensions contain a given permutation generalizing the binary tree insertion map on permutations (\cref{subsec:sweepingAlgorithm,subsec:insertionAlgorithm}), we describe the pipe dream congruence as the transitive closure of a local rewriting rule generalizing that of the sylvester congruence (\cref{subsec:rewritingRule}), and we describe a natural commutative diagram of lattice morphisms generalizing the connection between the recoil map and the binary tree insertion map on permutations and the canopy on binary trees (\cref{subsec:canopy}).

@@

In this section, we discuss four further topics on the pipe dream congruence.
We first describe two algorithms to construct the pipe dream~$\insertion{\pi}{\omega}$ of~$\acyclicPipeDreams(\omega)$ of which a given permutation~$\pi$ is a linear extension (\cref{subsec:sweepingAlgorithm,subsec:insertionAlgorithm}).
We then describe the pipe dream congruence~$\equiv_\omega$ as the transitive closure of a rewriting rule on permutations of~$[e, \omega]$ (\cref{subsec:rewritingRule}).
Finally, we describe the natural coarsening of the pipe dream congruence~$\equiv_\omega$ by the recoil congruence~$\cong_\omega$ (\cref{subsec:canopy}).

@@

To conclude, we generalize the notion of recoils of permutations and of canopy of a binary trees to show a natural commutative diagram of lattice homomorphisms.
We start with the easy generalization of recoils of a permutation.

@@

@@

>>>>

\vspace*{.1cm}

@@

We then explore in \cref{sec:furtherTopics} some natural further topics on pipe dreams.
We first present two algorithms to compute the unique acyclic pipe dream whose linear extensions contain a given permutation generalizing the binary tree insertion map on permutations (\cref{subsec:sweepingAlgorithm,subsec:insertionAlgorithm}).
We then describe the pipe dream congruence as the transitive closure of a local rewriting rule generalizing that of the sylvester congruence (\cref{subsec:rewritingRule}).
We then present a natural commutative diagram of lattice morphisms generalizing the connection between the recoil map and the binary tree insertion map on permutations and the canopy on binary trees (\cref{subsec:canopy}).
Finally, we discuss a specific family of pipe dreams in connection to the $\nu$-Tamari lattices (\cref{subsec:nuTamari}).

@@

In this section, we discuss five further topics on the pipe dream congruence.
We first present two algorithms to construct the pipe dream~$\insertion{\pi}{\omega}$ of~$\acyclicPipeDreams(\omega)$ of which a given permutation~$\pi$ is a linear extension (\cref{subsec:sweepingAlgorithm,subsec:insertionAlgorithm}).
We then describe the pipe dream congruence~$\equiv_\omega$ as the transitive closure of a rewriting rule on permutations of~$[e, \omega]$ (\cref{subsec:rewritingRule}).
We then present the natural coarsening of the pipe dream congruence~$\equiv_\omega$ by the recoil congruence~$\cong_\omega$ (\cref{subsec:canopy}).
Finally, we discuss a specific family of pipe dreams in connection to the $\nu$-Tamari lattices (\cref{subsec:nuTamari}).

@@

We now generalize the notion of recoils of permutations and of canopy of a binary trees to show a natural commutative diagram of lattice homomorphisms.
We start with the easy generalization of recoils of a permutation.

@@

\subsection{The acyclic property for $\nu$-Tamari lattices}
\label{subsec:nuTamari}

We close this section by stating an open problem that relates our results to the $\nu$-Tamari lattices introduced by L.-F.~Préville-Ratelle and X.~Viennot in~\cite{PrevilleRatelleViennot}. These posets are indexed by a lattice path $\nu$ consisting of a finite number of north~(N) and east~(E) steps, and coincide with the classical Tamari lattices when $\nu=(NE)^n$.

In~\cite{CeballosPadrolSarmiento}, it was shown that the $\nu$-Tamari lattice can be obtained as the increasing flip poset on pipe dreams~$\pipeDreams(\omega_\nu)$ for an explicit permutation $\omega_\nu$ associated to $\nu$. The permutations of the form~$\omega_\nu$ can be easily characterized as follows. We refer to~\cite{CeballosPadrolSarmiento} for details.

The \defn{Rothe diagram} of a permutation~$\omega$ is the set $\set{(\omega(j),i)}{i<j,\ \omega(i)>\omega(j)}$ in matrix notation.
A permutation $\omega$ is called \defn{dominant} if its Rothe diagram is the Ferrer's diagram of a partition located at the top left corner $(1,1)$. Equivalently, a permutation~$\omega$ is dominant if and only if it is 132-avoiding. 
For a permutation~$\omega = \omega_1 \dots \omega_n \in \fS_n$, we denote by~$0\omega$ the permutation~$0 \omega_1 \dots \omega_n$ of~$\{0, 1, \dots, n\}$, and we consider here pipe dreams with pipes indexed by~$\{0, 1, \dots, n\}$.
%For instance, for the reversed permutation~$\omega = n \dots 2 1$, the permutation~$0\omega$ is the permutation~$\rho_n$ considered in \cref{exm:Tamari1}.
%The permutations $\omega_\nu$ associated to lattice paths $\nu$ are those of the form~$1\oplus w$ for some dominant permutation $w$, see~\cite{CeballosPadrolSarmiento}.

\begin{openproblem}%[The acyclic property for $\nu$-Tamari lattices]
\label{prob:nuAcyclicProperty}
Show that if $\omega$ is a dominant permutation then all pipe dreams with permutation $0 \omega$ are acyclic. %In particular, the corresponding $\nu$-Tamari lattice is the increasing flip poset on acyclic pipe dreams~$\acyclicPipeDreams(1\oplus \omega)$.
\end{openproblem}

From \cref{exm:Tamari1}, we know that this property holds when $\omega=n\dots21$ is the reverse permutation. 
In this case, the increasing flip poset on~$\pipeDreams(0 \omega)$ is the classical Tamari lattice. Removing vertex $0$ from the contact graph of a pipe dream~$ P\in \pipeDreams(0 \omega)$ returns the binary tree corresponding to $P$ (with the leaves removed and the internal nodes labeled from $1$ to $n$).
The edges of the binary tree are oriented going away from the root of the tree.  
See \cref{fig:bijection}.
We have also checked this property computationally for all dominant permutations up to $n=8$.

Applying our Theorem~\ref{thm:pipeDreamQuotient}, we get the following consequence.

\begin{proposition}
If the statement of Open Problem~\ref{prob:nuAcyclicProperty} holds, then the $\nu$-Tamari lattice is a lattice quotient of the interval~$[e,\omega_\nu]$.
\end{proposition}

@@

\section*{Acknowledgments}

We are grateful to Florent Hivert for various discussions. We also thank Martin Rubey for his comments which led us to think about Open Problem~\ref{prob:nuAcyclicProperty}.
